---
title: "Report for PS 1"
author:
- Mark Becker, Erik Lillrank & Vilim Nedic
- Another Author
date: ""
output:
  pdf_document:
    keep_tex: yes
  df_print: kable
  html_document:
    df_print: paged
toc: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(FE)
library(psych)
data("DJ_w")
data("DJ_d")
```

#  Statistical Properties of Asset Returns

## A. Distributional properties of Dow Jones index returns

### Plot the series of log returns and compute descriptive statistics. Which distributional properties do you find? Do you find differences between daily and weekly data?

```{r PLOT1, echo=F}
ggplot(DJ_d, aes(y=r_Dow_Jones,x=1:nrow(DJ_d))) + geom_line() +
  labs(x='time horizon',y='daily log return')

ggplot(DJ_w, aes(y=r_close,x=1:nrow(DJ_w))) + geom_line() +
  labs(x='time horizon',y='weekly log return')

ggplot(DJ_d, aes(r_Dow_Jones)) + geom_density()
ggplot(DJ_w, aes(r_close)) + geom_density()
```

We can see that the densities look more skewed than the standard normal distribution, and more density close to the means.

```{r PLOT1, echo=F}
describe(DJ_d$r_Dow_Jones)
describe(DJ_w$r_close)
```
The mean of both series are 0. For the daily returns, there seems to be some skewness, as the absolute value of the minimum is a bit larger than the maximum. This is also the case for the weekly returns.

```{r PLOT2, echo=F}
ggplot(DJ_d, aes(r_Dow_Jones)) + geom_density()
ggplot(DJ_d, aes(r_Dow_Jones)) + geom_histogram(binwidth=.002)
ggplot(DJ_d) + geom_qq(aes(sample=r_Dow_Jones/sd(r_Dow_Jones))) +
  geom_abline(aes(intercept=0,slope=1),color="red")
```

# Question 2
## A)
```{r descr, echo=F}
describe(DJ_d$r_Dow_Jones)
describe(DJ_w$r_close)
```
For daily return data we see that we have 18839 observations. The time series are centered around mean 0 with standard deviation 0.01. It is slightly skewed with a statistic of -0.96 and kurtosis of 36.6. This indicates that the time series is distributed with a t distribution of XX degrees of freedom. 

For weekly returns we have 4686 observations instead with mean 0 and standard deviation 0.03. This time series is slightly more skewed with -1.09 but kurtosis is 15.06, also indicating t distributed returns with XX degrees of freedom. 
```{r tdist}
df = 3
tmp = ggplot(DJ_d, aes(sample=r_Dow_Jones/sd(r_Dow_Jones)*sqrt(df/(df-2))))
tmp + geom_qq(distribution=stats::qt, dparams=list(df=df)) + geom_abline(aes(intercept=0,slope=1),color="red")
#qqplot(rt(length(DJ_d$r_Dow_Jones),df=5),DJ_d$r_Dow_Jones)
```


```{r good, include=F}
vdata = DJ_d$r_Dow_Jones
#vdata = DJ_d$r_close
vdata = (vdata - mean(vdata))/sd(vdata)
ik = 20
grids = 1:ik/ik
```


```{r gooda}
vq = pnorm(vdata)
hist(vq)
ggplot(data=tibble(vq),aes(vq))+geom_histogram()
vn = NULL; for(val in grids) vn = c(vn,sum(vq <= val))
vn = c(vn[1],diff(vn))
test = sum((vn-length(vdata)/ik)**2/(length(vdata)/ik))
cat("test =",test," df =",ik-3," p-value =",1-pchisq(test,df=ik-3))
```


```{r goodb}
df = 3
ndata = vdata*sqrt(df/(df-2))
vq = pt(ndata,df=5); hist(vq)
vn = NULL; for(val in grids) vn = c(vn,sum(vq <= val))
vn = c(vn[1],diff(vn))
test = sum((vn-length(vdata)/ik)**2/(length(vdata)/ik))
cat("test =",test," df =",ik-3," p-value =",1-pchisq(test,df=ik-3))
```

```{r}
alpha = 0.1514736
sigma = 4.0013995
ndata = vdata * sqrt((1-alpha) + alpha*sigma**2)
vq = (1-alpha)*pnorm(ndata) + alpha*pnorm(ndata,sd=sigma); hist(vq)
vn = NULL; for(val in grids) vn = c(vn,sum(vq <= val))
vn = c(vn[1],diff(vn))
test = sum((vn-length(vdata)/ik)**2/(length(vdata)/ik))
cat("test =",test," df =",ik-3," p-value =",1-pchisq(test,df=ik-3))
```

Write a function for minimization to choose the optimal pair for $\alpha$ and $\sigma$.
```{r}
func <- function(vx){
  alpha = vx[1]
  sigma = vx[2]
  ndata = vdata*sqrt((1-alpha) + alpha*sigma**2)

  vq = (1-alpha)*pnorm(ndata) + alpha*pnorm(ndata,sd=sigma)
  vn = NULL; for(val in grids) vn = c(vn,sum(vq <= val))
  vn = c(vn[1],diff(vn))
  return(sum((vn-length(vdata)/ik)**2/(length(vdata)/ik)))
}

func(c(0.15,4))
```


```{r}
optim(par=c(0.15,4),fn=func,method="BFGS")
optim(par=c(0.15,4),fn=func,method="L-BFGS-B",lower=c(0,0),upper=c(1,10)) 
```

```{r, cache=T}
atmp = seq(from=0.05, to=.95, length.out=40)
stmp = seq(from=1, to=10, length.out=50)
tmp = expand.grid(atmp,stmp)
res = apply(tmp,1,func)
res = t(matrix(res, nrow=length(atmp)))

library(plotly)
ret = plot_ly(x=atmp, y=stmp, z=res) %>% add_surface() %>%
  layout(scene=list(xaxis=list(title='alpha'), yaxis=list(title='sigma'), zaxis=list(title='function')))
ret
```

## B Dynamical properties of financial return series

```{r setupb, include=F}
data = index_d
```

## 1. Generate log returns. Compute the empirical autocorrelations and partial autocorrelations. Do you find evidence for significant autocorrelations in the individual index return series?

```{r}
lret = apply(log(data),2,diff)
summary(lret)
matplot(lret,type='l',ylab='log returns',xlab='time horizon')

sum(is.na(lret[,'FRCAC40']))
```

## 2. Compute the Ljung-Box test with respect to 10, 50 and 100 lags. By using the 0.01, 0.05 and 0.1 quantile of the χ2-distribution with suitable degrees of freedom, interpret your results.

```{r}
LB <- function(vx,lag,ip){
  tmp = acf(vx,lag.max=lag,plot=F)$acf
  tmp = tmp[2:(lag+1)]**2
  test = sum(tmp/(length(vx)-1:lag))*length(vx)*(length(vx)+2)
  return(list(test=test, pval=1-pchisq(test,df=lag-ip)))
}

tmp = 6
LB(vx=lret[!is.na(lret[,tmp]),tmp],lag=100,ip=0)
```

