---
title: "Report for PS4"
author:
- Mark Becker, Erik Lillrank & Vilim Nedic
date: ''
output:
  pdf_document:
    keep_tex: true
  df_print: kable
  html_document:
    df_print: paged
  word_document: default
header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{multirow}
  - \usepackage{caption}
  - \usepackage{subcaption}
  - \usepackage{adjustbox}
  - \usepackage{hyperref}
toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(FE)
library(knitr)
library(kableExtra)
library(dplyr)
library(ggplot2)
library(psych)
library(systemfit)
data("portfolio_m")
```

# A. Factor models with portfolios as factors

In this exercise you have to estimate and to test a multi-factor model with portfolios as factors. Ise the data set \texttt{portfolio_m}

## 1. Compute factor portfolios based on principal components using the correlation matrix of returns.

### (a) Create the factor portfolios based on the first 5 principal components using the covariance matrix of the 100 size-book-to-market portfolios and save them. Evaluate how much of the variation of these 100 portfolios can be explained by the principal components.

```{r}
pca_res_100 <- PCA(as.matrix(portfolio_m[,25:124]), 5)

pc_scores_100 <- pca_res_100$factors
colnames(pc_scores_100) <- paste0("PC", 1:5)

proportion_variance <- pca_res_100$explained
cumulative_variance <- cumsum(proportion_variance)

variance_table <- rbind(
  `Proportion of Variance` = proportion_variance,
  `Cumulative Proportion` = cumulative_variance
)

colnames(variance_table) <- paste0("PC", 1:5)

kable(variance_table, caption="The variation of the 100 portfolios that can be explained by the first 5 principal components. In percentages.")
```
Approximately $77\%$ of the variance in the 100 size-book-to-market portfolios can be explained by the first 5 principal components.

### (b) Create the factor portfolios based on the first 5 principal components using the covariance matrix of the 20 size and book-to-market portfolios and save them as well. Evaluate how much of the variation of these 20 portfolios can be explained by the principal components.

```{r}
pca_res_20 <- PCA(as.matrix(portfolio_m[,5:24]), 5)  

pc_scores_20 <- pca_res_20$factors
colnames(pc_scores_20) <- paste0("PC", 1:5)

proportion_variance <- pca_res_20$explained
cumulative_variance <- cumsum(proportion_variance)

variance_table <- rbind(
  `Proportion of Variance` = proportion_variance,
  `Cumulative Proportion` = cumulative_variance
)

colnames(variance_table) <- paste0("PC", 1:5)

kable(variance_table, caption="The variation of the 20 portfolios that can be explained by the first 5 principal components. In percentages.")
```

Approximately $97\%$ of the variance in the 20 size and book-to-market portfolios can be explained by the first 5 principal components.

### (c) Evaluate the distributional properties of the individual factor portfolios. What do you find?

We will start by presenting some key descriptive statistics of the factor portfolios.

```{r}
descr_100 <- describe(pc_scores_100)[,c("mean", "sd", "skew", "kurtosis", "min", "max")]

rownames(descr_100) <- paste0("PC", 1:5)

kable(descr_100, digits = 3, caption = "Descriptive statistics of the factor portfolios for the 100 size-book-to-market portfolios.")
```
In the table above you see the descriptive statistics for the factors of the 100 size-book-to-market portfolios. The first factor portfolio has large negative average return compared the other factors which have average returns closer to  zero. The standard deviations of the factor portfolios is decreasing in the order PC1 to PC5, this is because the first principal component will by design always capture the most of the variability in the data.

The first, third and fourth factors are heavily skewed, with PC1 and PC3 having negative skewness and PC4 having positive skewness. The negative skewness indicates that these portfolios are influenced by large negative returns, and the opposite for the positive skewness. All factor portfolios have high kurtosis indicating fat tails.

```{r}
descr_20 <- describe(pc_scores_20)[,c("mean", "sd", "skew", "kurtosis", "min", "max")]

rownames(descr_20) <- paste0("PC", 1:5)

kable(descr_20, digits = 3, caption = "Descriptive statistics of the factor portfolios for the 20 size and book-to-market portfolios.")
```
In the table above we see the descriptive statistics for the factor portfolios of the 20 size and book-to-market portfolios. Here we see similar patterns as for the factor portfolios of the 100 size-book-to-market portfolios. The average returns are negative for the first factor portfolio and close to zero for the other. PC1 captures most of the variation in the original data. The factor portfolios differ in their skewness and kurtosis, PC1 and PC2 are moderately skewed while the other factors are only slightly skewed, and all factor portfolios have heavy tails.

### (d) How strongly are the two sets of factor portfolios (created under (a) and (b)) correlated with each other? How strongly are the individual factor portfolios correlated with the market portfolio. What can you conclude?

```{r}
correlation_matrix <- cor(pc_scores_100, pc_scores_20)
kable(round(correlation_matrix,2))
```
The rows corresponds to the factor portfolios of the 100 size-book-to-market portfolios, and the columns to the factor portfolios of the 20 size and book-to-market portfolios. The first factors are very highly correlated with 0.99 correlation. There are also high correlations between the second factors, as well as between the third factors. PC2 and PC3 are also highly correlated between the two sets of factor portfolios.

```{r}
correlations_with_market <- sapply(colnames(pc_scores_100), function(factor) {
  cor(pc_scores_100[,factor], portfolio_m[,3])
})

kable(round(correlations_with_market,2), caption = "Correlation of each factor portfolio (of the 100 size-book-to-market) with the market portfolio:")

correlations_with_market_20 <- sapply(colnames(pc_scores_20), function(factor) {
  cor(pc_scores_20[,factor], portfolio_m[,3])
})

kable(round(correlations_with_market_20,2), caption = "Correlation of each factor portfolio (of the 20 size and book-to-market) with the market portfolio:")
```

The first factor for each of the sets are most highly correlated with the market, with the second factor having the second highest correlation with the market, and so forth. The first factors likely captures the market-wide risks and movements. Factors 3, 4, and 5 are not correlated with the market indicating that the variability these factors capture is not market wide.

## 2. Compute the excess returns for all portfolios. In the following you have to estimate a multi-factor model where you include either the principal components constructed based on the 100 size-book-to-market portfolios or the principal components constructed based on the 20 size and book-to-market portfolios as factors. Estimate the model for individual size-book-to-market portfolios in the 1st, 5th and 10th size and book-to-market decile.

```{r}
mR = as.matrix(portfolio_m[,25:124])
Rf = as.matrix(portfolio_m[,'Tbill'])
Rm = as.matrix(portfolio_m[,'Market'])
mZ = sweep(mR,1,Rf) #matrix of excess returns for the portfolios
vZm = Rm - Rf #market excess returns

pca_res_100 <- PCA(as.matrix(portfolio_m[,25:124]), 5) #5 Principal components based on the 100 sbtm portfolios
pc_scores_100 <- pca_res_100$factors
pc_scores_excess_100 <- sweep(pc_scores_100, 1, Rf) #excess returns of factor portfolios

portfolio <- c(1,5,10,41,45,50,91,95,100)
factors <- c(1,2,3,4,5)

mR2 <- matrix(NA,nrow=length(portfolio),ncol=length(factors))
mBIC <- matrix(NA,nrow=length(portfolio),ncol=length(factors))

for (i in 1:length(portfolio)){
  for (j in 1:length(factors)){
    model <- lm(mZ[, portfolio[i]] ~ pc_scores_excess_100[, 1:j])
    mR2[i, j] <- summary(model)$r.squared
    mBIC[i, j] <- BIC(model)
  }
}
```


### (a) How much of the time series variations of the portfolio returns can be explained by the individual principal components? How strong does the goodness-of-fit increase when we include more than one factor? What is the optimal number of factors according to the Bayes/Schwarz information criterion?

```{r}
R2_table <- cbind(colnames(mZ)[c(1,5,10,41,45,50,91,95,100)],round(mR2,2))
colnames(R2_table) <- c("Portfolio", "1 Factor", "2 Factors", "3 Factors", "4 Factors", "5 Factors")
kable(R2_table, caption = "R-squared values for 1-5 factor model")

BIC_table <- cbind(colnames(mZ)[c(1,5,10,41,45,50,91,95,100)],round(mBIC,0))
colnames(BIC_table) <- c("Portfolio", "1 Factor", "2 Factors", "3 Factors", "4 Factors", "5 Factors")
kable(BIC_table, caption = "BIC values for 1-5 factor models")
```

BIC suggests 5,5,5,4,4,5,5,4,5 factors for the respective portfolios. The difference is very small for the portfolios where it suggests 4 instead of 5.

### (b) Do you find differences in the explanatory power of the two sets of principal components?

For simplicity lets assume 5 factor model is the best.

```{r}
pca_res_20 <- PCA(as.matrix(portfolio_m[,5:24]), 5) #20 size book, 5 factors
pc_scores_20 <- pca_res_20$factors
pc_scores_excess_20 <- sweep(pc_scores_20, 1, Rf)

pca_res_100 <- PCA(as.matrix(portfolio_m[,25:124]), 5) #100 size book, 5 factors
pc_scores_100 <- pca_res_100$factors
pc_scores_excess_100 <- sweep(pc_scores_100, 1, Rf)


portfolio <- c(1,5,10,41,45,50,91,95,100)

mR2_20_100 <- matrix(NA, nrow = length(portfolio), ncol = 2)

for (i in 1:length(portfolio)){
    model_20 <- lm(mZ[, portfolio[i]] ~ pc_scores_excess_20)
    mR2_20_100[i,1] <- summary(model_20)$r.squared
    model_100 <- lm(mZ[, portfolio[i]] ~ pc_scores_excess_100)
    mR2_20_100[i,2] <- summary(model_100)$r.squared
}


R2_comparison <- data.frame(
  Portfolio = colnames(mZ)[portfolio],
  `PCA based on 20 MEx/MEBEx portfolios` = round(mR2_20_100[, 1], 2),
  `PCA based on 100 Rx(x)y(y) portfolios` = round(mR2_20_100[, 2], 2)
)

kable(R2_comparison, caption = "R-squared values for 5 factor models")
```

### (c) Evaluate the explanatory power of the individual principal components and compare it to that provided by the market risk premium solely. Does the market risk premium have any explanatory power beyond the principal components? What is about the significance of the market risk premium in such a regression? What can you conclude?

```{r}
mR = as.matrix(portfolio_m[,25:124])
Rf = as.matrix(portfolio_m[,'Tbill'])
Rm = as.matrix(portfolio_m[,'Market'])
mZ = sweep(mR,1,Rf) #matrix of excess returns for all portfolios
vZm = Rm - Rf #matrix of excess returns for the market portfolio

portfolio <- c(1,5,10,41,45,50,91,95,100)

vR2_market <- matrix(NA,nrow=length(portfolio),1)
vtstat_market <- matrix(NA,nrow=length(portfolio),1)
vpval_market <- matrix(NA,nrow=length(portfolio),1)



for (i in 1:length(portfolio)){
    model <- lm(mZ[, portfolio[i]] ~ vZm)
    vR2_market[i, 1] <- summary(model)$r.squared
    vtstat_market[i,1] <-  summary(model)$coefficients[2, "t value"]
    vpval_market[i,1] <-  summary(model)$coefficients[2, "Pr(>|t|)"]
}

R2_table <- cbind(colnames(mZ)[c(1,5,10,41,45,50,91,95,100)],round(mR2,2))
R2_table_final <- cbind(R2_table,round(vR2_market,2))
colnames(R2_table_final) <- c("Portfolio", "1 Factor", "2 Factors", "3 Factors", "4 Factors", "5 Factors", "1 Factor - Market risk premium")
kable(R2_table_final, caption = "R-squared values for 1-5 factor model and 1 Factor market model")

# Combine results into a data frame
market_stats <- data.frame(
  Portfolio = colnames(mZ)[portfolio],
  `T-statistic` = round(vtstat_market, 2),
  `P-value` = round(vpval_market, 4)
)

# Display the results as a table
kable(market_stats, caption = "Significance of the market risk premium")

```

It is significant at the 1\% level for all portfolios.

### (d) For which type of portfolio do you find the best goodness-of-fit? For which type of portfolio do you find the worst goodness-of-fit? Interpret your findings.

We can see in the table above that the 1 factor model has the lowest R squareds. The best...

## 3. Use the principal components constructed based on the 20 size and book- to-market portfolios in a multi-factor model and estimate it as a system for all 100 size-book-to- market portfolios. Test jointly for exact factor pricing in all equations. Furthermore, test for the joint significance of the factors. Use an F-test as well as a likelihood ratio test. What do you find?

To answer this question, since we have the PCA for 20, we want to estimate a system of equations. The OLS estimator for this becomes as following:

$$
\hat{\alpha}=\bar{Z}-\hat{\beta}\hat{\mu}_K, \\
\text{where }
\bar{Z}=\frac{1}{T} \sum_{t=1}^{T}Z_t, \text{ }\hat{\mu}_k=\frac{1}{T}\sum_{t=1}^{T}Z_{Kt},\\
\hat{\beta}= [\sum_{t=1}^{T}(Z_t-\bar{Z})(Z_{Kt}-\hat{\mu}_k)' ][\sum_{t=1}^{T} (Z_t-\hat{\mu}_K)(Z_{Kt}-\hat{\mu}_k)' ]
$$
The covariance matrix of the error terms are estimated by maximum likelihood: 
$$
\hat{\Sigma}=\frac{1}{T} \sum_{t=1}^{T} (Z_t-\hat{\alpha}-\hat{\beta}Z_{Kt})(Z_t-\hat{\alpha}-\hat{\beta}Z_{Kt})'
$$
The F-test proposed by Gibbons/Ross/Shanken (1989) is:
$$
J=\frac{T-N-K}{K}[1+\hat{\mu}_K' \hat{\Omega}_K^{-1} \hat{\mu}_K]^{-1} \hat{\alpha}' \hat{\Sigma}^{-1} \hat{\alpha} \sim F_{N, T-N-K}
$$
And the adjusted likelihood ratio test is: 

$$
LR=(T-\frac{N}{2}-K-1) [\text{log }|\hat{\Sigma}_R|-\text{log }|\hat{\Sigma}|] \sim \chi_N^2
$$
where the restricted model is with intercept = 0 and the unrestricted model is the full model. 

Since we have stock portfolios as factors we aim to estimate a SUR. 
F-test based on Gibbons/Ross/Shanken (1989). 

To estimate a SUR model, the following equations are needed 
```{r}
library(systemfit)
mR = as.matrix(portfolio_m[,25:124])
Rf = as.matrix(portfolio_m[,'Tbill'])
mZ = sweep(mR,1,Rf) #matrix of excess returns for the portfolios

pca_res_20 <- PCA(as.matrix(portfolio_m[,5:24]), 5) #20 size book, 5 factors
pc_scores_20 <- pca_res_20$factors
mZk <- sweep(pc_scores_20, 1, Rf)

T <- nrow(mZ)

# Step 1: Compute means
Z_bar <- colMeans(mZ)
mu_k <- colMeans(mZk)

# Step 2: Estimate B
cov_Z_Zk <- crossprod(mZ - matrix(Z_bar, nrow = T, ncol = ncol(mZ), byrow = TRUE), 
                      mZk - matrix(mu_k, nrow = T, ncol = ncol(mZk), byrow = TRUE))
cov_Zk <- crossprod(mZk - matrix(mu_k, nrow = T, ncol = ncol(mZk), byrow = TRUE))
B_hat <- cov_Z_Zk %*% solve(cov_Zk)

# Step 3: Estimate alpha
alpha_hat <- Z_bar - B_hat %*% mu_k

# Step 4: Estimate Sigma
residuals <- Z - (matrix(alpha_hat, nrow = T, ncol = ncol(Z), byrow = TRUE) + Z_k %*% t(B_hat))
Sigma_hat <- crossprod(residuals) / T


```



## 4. Test for joint significance of the market risk premium in all equations. Does it have explanatory power beyond the principal components? Compare the goodness-of-fit of alternative factor specifications using information criteria. Which model provides the highest goodness-of-fit for the complete system?




# B. Fama-French regressions
Use the data set portfolio_m and compute excess returns. In the following you have to mimic the study by Fama and French (1996) and have to investigate whether the Fama-French factors capture size and book-to-market effects.


## 1. Construct the HML and SMB factors according to Fama and French (1993) based on the return differentials of small-cap and large-cap portfolios as well as high-book-to-market and small-book-to-market portfolios.

In order to construct the HML and SMB factors, we take the mean of the three highest BtM-portfolios minus the mean of the three smallest BtM portfolios for the HML, and for SMB we take the mean of the smallest three portfolios minus the mean of the three largest portfolios.

```{r}
HML <- rowMeans(portfolio_m[,22:24])-rowMeans(portfolio_m[,15:17])
SMB <- rowMeans(portfolio_m[,5:7])-rowMeans(portfolio_m[,12:14])

portfolio_factors <- portfolio_m %>%
  mutate(HML = HML, SMB = SMB)

head(portfolio_factors[,c(1:4, 125:126)])
```

## 2. Estimate the resulting Fama-French three-factor model based on individual size-book-to-market portfolios in the 1st, 5th and 10th size and book-to-market decile.

```{r}
cols <- c("R11", "R15", "R110", "R51", "R55", "R510", "R101", "R105", "R1010")

results <- data.frame(
  Portfolio = character(),
  Alpha = numeric(),
  Beta = numeric(),
  s = numeric(),
  h = numeric(),
  R2 = numeric(),
  R2_comp = numeric(),
  StdErrAlpha = numeric(),
  StdErrBeta = numeric(),
  StdErr_s = numeric(),
  StdErr_h = numeric(), 
  PValueAlpha = numeric(),
  PValueBeta = numeric(),
  PValue_s = numeric(),
  Pvalue_h = numeric(),
  stringsAsFactors = FALSE
)

for (i in cols){
  ER <- portfolio_factors[[i]]-portfolio_factors$Tbill
  market_excess <- portfolio_factors$Market-portfolio_factors$Tbill
  capm_model <- lm(ER ~ market_excess+portfolio_factors$SMB+portfolio_factors$HML)
  
  capm_comp <- lm(ER ~ market_excess)
  
  alpha <- coef(capm_model)["(Intercept)"]
  beta <- coef(capm_model)["market_excess"]
  s <- coef(capm_model)["portfolio_factors$SMB"]
  h <- coef(capm_model)["portfolio_factors$HML"]
  r_squared <- summary(capm_model)$r.squared
  std_err_alpha <- coef(summary(capm_model))["(Intercept)", "Std. Error"]
  std_err_beta <- coef(summary(capm_model))["market_excess", "Std. Error"]
  std_err_s <- coef(summary(capm_model))["portfolio_factors$SMB", "Std. Error"]
  std_err_h <- coef(summary(capm_model))["portfolio_factors$HML", "Std. Error"]
  p_value_alpha <- coef(summary(capm_model))["(Intercept)", "Pr(>|t|)"]
  p_value_beta <- coef(summary(capm_model))["market_excess", "Pr(>|t|)"]
  p_value_s <- coef(summary(capm_model))["portfolio_factors$SMB", "Pr(>|t|)"]
  p_value_h <- coef(summary(capm_model))["portfolio_factors$HML", "Pr(>|t|)"]
  
  r_squared_comp <- summary(capm_comp)$r.squared
  
  results <- rbind(
    results,
    data.frame(
      Portfolio = i,
      Alpha = alpha,
      Beta = beta,
      s=s,
      h=h,
      R2 = r_squared,
      R2_comp = r_squared_comp, 
      StdErrAlpha = std_err_alpha,
      StdErrBeta = std_err_beta,
      StdErr_s = std_err_s,
      StdErr_h = std_err_h, 
      PValueAlpha = p_value_alpha,
      PValueBeta = p_value_beta,
      PValue_s = p_value_s,
      Pvalue_h = p_value_h
    )
  )
}

rownames(results) <- NULL
```

### (a) Evaluate the goodness-of-fit of the model for the individual portfolios. How much explanatory power do the HML and SMB factors have beyond the market portfolio?

```{r}
kable(results[,c(1,6:7)], digits = 3, caption = "Comparison of R2 between the estimated three factor and CAPM model for the portfolios", label = "tab:3factor")
```
In the table above you see a comparison of the $R^2$ between the Fama and French three factor model and the regular CAPM. What we see is that the $R^2$ increases for all portfolios when including SMB and HML in the model, indicating that HML and SMB factors have explanatory power beyond the market portfolio. The increase in explanatory power is largest for the smaller portfolios (R11, R15, and R110), with the smallest increase for the largest portfolios (R101, R105, and R1010). We also see that the change is larger for the portfolios with higher book-to-market deciles (R110, R510, and R110) compared to the lower deciles. 

Based on this it seems that the explanatory power increases most for the more volatile portfolios, indicating that there is variation explained by the factors rather than the market. This reasonable as the lower book-to-market and larger size portfolios are less volatile and follow the market more closely.

### (b) Compare the estimated coefficients associated with the HML and SMB factors for portfolios in different size and book-to-market deciles. Do you find differences? Interpret your results.

```{r}
kable(results[,c(1,4:5)], digits = 3, caption = "Comparison of estimated coefficients of the estimated three factor model between the portfolios", label = "tab:3factor_coeff")
```
Looking at the estimated coefficients above, we see that estimated $s_i$ is positive for the small and medium size portfolios (R1x and R5x), and negative for the larger size portfolios (R10x). The Estimated $s_i$ is also largest for the smallest size portfolios, and smaller for the medium and large size portfolios. The estimated $h_i$ are negative for the low book-to-market portfolios and positive for the high and medium (except R15). There is no clear correlation between the magnitude of the h coefficient and book-to-market.

Based on this we conclude that SMB more greatly explain the variation in the smaller size portfolios and that there is a negative correlation between SMB and larger size portfolios. We also conclude that HML seems to be positively correlated with high book-to-market portfolios and negatively correlated with low book-to-market portfolios.

### (c) Augment your model by the first two principal components as constructed in Problem A, Part 2. Do the principle components have explanatory power beyond the Fama- French factors? Do the Fama-French factors have explanatory power beyond the principal components? Interpret your findings.

```{r}
colnames(pc_scores_excess) <- paste0("Excess_PC", 1:5)
pc_scores_excess <- as.data.frame(pc_scores_excess)
cols <- c("R11", "R15", "R110", "R51", "R55", "R510", "R101", "R105", "R1010")

resultsb3c <- data.frame(
  Portfolio = character(),
  Alpha = numeric(),
  Beta = numeric(),
  s = numeric(),
  h = numeric(),
  PC1 = numeric(),
  PC2 = numeric(),
  R2 = numeric(),
  R2_comp = numeric(),
  StdErrAlpha = numeric(),
  StdErrBeta = numeric(),
  StdErr_s = numeric(),
  StdErr_h = numeric(), 
  StdErr_PC1 = numeric(),
  StdErr_PC2 = numeric(),
  PValueAlpha = numeric(),
  PValueBeta = numeric(),
  PValue_s = numeric(),
  Pvalue_h = numeric(),
  PValue_PC1 = numeric(),
  PValue_PC2 = numeric(),
  stringsAsFactors = FALSE
)

for (i in cols){
  ER <- portfolio_factors[[i]]-portfolio_factors$Tbill
  market_excess <- portfolio_factors$Market-portfolio_factors$Tbill
  capm_model <- lm(ER ~ market_excess + portfolio_factors$SMB + portfolio_factors$HML + pc_scores_excess$Excess_PC1 + pc_scores_excess$Excess_PC2)
  
  comp <- lm(ER ~market_excess + pc_scores_excess$Excess_PC1 + pc_scores_excess$Excess_PC2)
  
  
  alpha <- coef(capm_model)["(Intercept)"]
  beta <- coef(capm_model)["market_excess"]
  s <- coef(capm_model)["portfolio_factors$SMB"]
  h <- coef(capm_model)["portfolio_factors$HML"]
  PC1 <- coef(capm_model)["pc_scores_excess$Excess_PC1"]
  PC2 <- coef(capm_model)["pc_scores_excess$Excess_PC2"]
  r_squared <- summary(capm_model)$r.squared
  r_squared_comp <- summary(comp)$r.squared
  std_err_alpha <- coef(summary(capm_model))["(Intercept)", "Std. Error"]
  std_err_beta <- coef(summary(capm_model))["market_excess", "Std. Error"]
  std_err_s <- coef(summary(capm_model))["portfolio_factors$SMB", "Std. Error"]
  std_err_h <- coef(summary(capm_model))["portfolio_factors$HML", "Std. Error"]
  std_err_PC1 <- coef(summary(capm_model))["pc_scores_excess$Excess_PC1", "Std. Error"]
  std_err_PC2 <- coef(summary(capm_model))["pc_scores_excess$Excess_PC2", "Std. Error"]
  p_value_alpha <- coef(summary(capm_model))["(Intercept)", "Pr(>|t|)"]
  p_value_beta <- coef(summary(capm_model))["market_excess", "Pr(>|t|)"]
  p_value_s <- coef(summary(capm_model))["portfolio_factors$SMB", "Pr(>|t|)"]
  p_value_h <- coef(summary(capm_model))["portfolio_factors$HML", "Pr(>|t|)"]
  p_value_PC1 <- coef(summary(capm_model))["pc_scores_excess$Excess_PC1", "Pr(>|t|)"]
  p_value_PC2 <- coef(summary(capm_model))["pc_scores_excess$Excess_PC2", "Pr(>|t|)"]
  
  
  resultsb3c <- rbind(
    resultsb3c,
    data.frame(
      Portfolio = i,
      Alpha = alpha,
      Beta = beta,
      s=s,
      h=h,
      PC1 = PC1,
      PC2 = PC2, 
      R2 = r_squared,
      R2_comp = r_squared_comp,
      StdErrAlpha = std_err_alpha,
      StdErrBeta = std_err_beta,
      StdErr_s = std_err_s,
      StdErr_h = std_err_h, 
      StdErr_PC1 = std_err_PC1,
      StdErr_PC2 = std_err_PC2,
      PValueAlpha = p_value_alpha,
      PValueBeta = p_value_beta,
      PValue_s = p_value_s,
      Pvalue_h = p_value_h,
      PValue_PC1 = p_value_PC1,
      PValue_PC2 = p_value_PC2
    )
  )
}

rownames(resultsb3c) <- NULL
kable(cbind(resultsb3c[,c(1,8:9)],results[,6]), digits = 3, caption = "Comparison of R2 between the estimated three factor and CAPM model for the portfolios", label = "tab:3factor_pca_r2", col.names =  c("Portfolio", "R2 (three factor + PCA)", "R2 (PCA)", "R2 (three factors)"))
```
Above you see a comparison in $R^2$ between three models: The augmented model with HML and SMB factors as well as the first two principal components, the model with only principal components, and a model with only the Fama-French factors. What we see is that including the PC's increases the goodness-of-fit compared to the three-factor model, meaning that the PC's have explanatory power beyond the Fama-French factors. However, there is only small differences in $R^2$ between this augmented model and the PC's model with the first two principal components, this indicates that the Fama-French factors have very little or no explanatory power beyond the principal components.

## 3. Estimate the Fama-French model for all 100 size-book-to-market portfolios as a system.

```{r}
# Pre-process data
excess_portfolio_factors <- cbind(portfolio_factors, PC1=pc_scores_excess$Excess_PC1,  PC2=pc_scores_excess$Excess_PC2)
excess_portfolio_factors <- excess_portfolio_factors %>%
  mutate(across(c(3, 25:124), ~ . - Tbill))

# Set factors
indep <- c("Market", "SMB", "HML", "PC1", "PC2")

Z <- as.matrix(excess_portfolio_factors[, 25:124])
Z_k <- as.matrix(excess_portfolio_factors[, indep])

T <- nrow(Z)

# Step 1: Compute means
Z_bar <- colMeans(Z)
mu_k <- colMeans(Z_k)

# Step 2: Estimate B
cov_Z_Zk <- crossprod(Z - matrix(Z_bar, nrow = T, ncol = ncol(Z), byrow = TRUE), 
                      Z_k - matrix(mu_k, nrow = T, ncol = ncol(Z_k), byrow = TRUE))
cov_Zk <- crossprod(Z_k - matrix(mu_k, nrow = T, ncol = ncol(Z_k), byrow = TRUE))
B_hat <- cov_Z_Zk %*% solve(cov_Zk)

# Step 3: Estimate alpha
alpha_hat <- Z_bar - B_hat %*% mu_k

# Step 4: Estimate Sigma
residuals <- Z - (matrix(alpha_hat, nrow = T, ncol = ncol(Z), byrow = TRUE) + Z_k %*% t(B_hat))
Sigma_hat <- crossprod(residuals) / T


```

### (a) Test jointly for exact factor pricing in all equations.

We will test $H_0:\alpha=0$ using the F-test based on Gibbons/Ross/Shanken (1989):

```{r}
N <- length(alphas)
K <- 3

Omega <- cov(Z_k)

GRS_F <- (T - N - K) / N * 
         (t(alpha_hat) %*% solve(Sigma_hat) %*% alpha_hat) /
         (1 + t(mu_k) %*% solve(Omega) %*% (mu_k))

GRS_p_value <- pf(GRS_F, df1 = N, df2 = T - N - K, lower.tail = FALSE)

results <- data.frame(
  Statistic = c("GRS F-statistic:", "p-value:"),
  Value = c(GRS_F, GRS_p_value)
)

kable(results, rownames = c("GRS F-statistic:", "p-value:"),
      caption = "F-test for exact factor pricing", digits = 3)
```
Exact factor pricing implies that model explain all of the variation in excess returns, if this holds true all $\alpha$ should be zero. Based on the above F-test we reject this null hypothesis and come to the conclusion that the factor pricing model does not perfectly explain portfolio returns.

### (b) Test jointly for the significance of the HML and SMB factors as well as the principal components. What do you find? Using information criteria find the model providing the best goodness-of-fit.

```{r}
restricted_eq <- lapply(eq, function(eq) update(eq, ~ . - SMB - HML))
restricted_model <- systemfit(restricted_eq, method = "SUR", data = excess_portfolio_factors)

LR <- 2 * (logLik(sur_model) - logLik(restricted_model))
df <- 2  
p_value <- pchisq(LR, df = df, lower.tail = FALSE)


```

### (c) Open the data set factors_m containing the original HML and SMB factors as constructed by Fama and French (1993). Repeat the analysis above. Are your results robust regarding the exact construction of the factors?