---
title: "Report for PS4"
author:
- Mark Becker, Erik Lillrank & Vilim Nedic
date: ''
output:
  pdf_document:
    keep_tex: true
  df_print: kable
  html_document:
    df_print: paged
  word_document: default
header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{multirow}
  - \usepackage{caption}
  - \usepackage{subcaption}
  - \usepackage{adjustbox}
  - \usepackage{hyperref}
toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(FE)
library(knitr)
library(kableExtra)
library(dplyr)
library(ggplot2)
library(psych)
library(systemfit)
library(msos)
data("portfolio_m")
```

# A. Factor models with portfolios as factors

In this exercise you have to estimate and to test a multi-factor model with portfolios as factors. Is the data set \texttt{portfolio_m}

## 1. Compute factor portfolios based on principal components using the correlation matrix of returns.

### (a) Create the factor portfolios based on the first 5 principal components using the covariance matrix of the 100 size-book-to-market portfolios and save them. Evaluate how much of the variation of these 100 portfolios can be explained by the principal components.

```{r}
pca_res_100 <- PCA(as.matrix(portfolio_m[,25:124]), 5)

pc_scores_100 <- pca_res_100$factors
colnames(pc_scores_100) <- paste0("PC", 1:5)

proportion_variance <- pca_res_100$explained
cumulative_variance <- cumsum(proportion_variance)

variance_table <- rbind(
  `Proportion of Variance` = proportion_variance,
  `Cumulative Proportion` = cumulative_variance
)

colnames(variance_table) <- paste0("PC", 1:5)

kable(variance_table, caption="The variation of the 100 portfolios that can be explained by the first 5 principal components. In percentages.")
```
Approximately $77\%$ of the variance in the 100 size-book-to-market portfolios can be explained by the first 5 principal components.

### (b) Create the factor portfolios based on the first 5 principal components using the covariance matrix of the 20 size and book-to-market portfolios and save them as well. Evaluate how much of the variation of these 20 portfolios can be explained by the principal components.

```{r}
pca_res_20 <- PCA(as.matrix(portfolio_m[,5:24]), 5)  

pc_scores_20 <- pca_res_20$factors
colnames(pc_scores_20) <- paste0("PC", 1:5)

proportion_variance <- pca_res_20$explained
cumulative_variance <- cumsum(proportion_variance)

variance_table <- rbind(
  `Proportion of Variance` = proportion_variance,
  `Cumulative Proportion` = cumulative_variance
)

colnames(variance_table) <- paste0("PC", 1:5)

kable(variance_table, caption="The variation of the 20 portfolios that can be explained by the first 5 principal components. In percentages.")
```

Approximately $97\%$ of the variance in the 20 size and book-to-market portfolios can be explained by the first 5 principal components.

### (c) Evaluate the distributional properties of the individual factor portfolios. What do you find?

We will start by presenting some key descriptive statistics of the factor portfolios.

```{r}
descr_100 <- describe(pc_scores_100)[,c("mean", "sd", "skew", "kurtosis", "min", "max")]

rownames(descr_100) <- paste0("PC", 1:5)

kable(descr_100, digits = 3, caption = "Descriptive statistics of the factor portfolios for the 100 size-book-to-market portfolios.")
```
In the table above you see the descriptive statistics for the factors of the 100 size-book-to-market portfolios. The first factor portfolio has large negative average return compared the other factors which have average returns closer to  zero. The standard deviations of the factor portfolios is decreasing in the order PC1 to PC5, this is because the first principal component will by design always capture the most of the variability in the data.

The first, third and fourth factors are heavily skewed, with PC1 and PC3 having negative skewness and PC4 having positive skewness. The negative skewness indicates that these portfolios are influenced by large negative returns, and the opposite for the positive skewness. All factor portfolios have high kurtosis indicating fat tails.

```{r}
descr_20 <- describe(pc_scores_20)[,c("mean", "sd", "skew", "kurtosis", "min", "max")]

rownames(descr_20) <- paste0("PC", 1:5)

kable(descr_20, digits = 3, caption = "Descriptive statistics of the factor portfolios for the 20 size and book-to-market portfolios.")
```
In the table above we see the descriptive statistics for the factor portfolios of the 20 size and book-to-market portfolios. Here we see similar patterns as for the factor portfolios of the 100 size-book-to-market portfolios. The average returns are negative for the first factor portfolio and close to zero for the other. PC1 captures most of the variation in the original data. The factor portfolios differ in their skewness and kurtosis, PC1 and PC2 are moderately skewed while the other factors are only slightly skewed, and all factor portfolios have heavy tails.

### (d) How strongly are the two sets of factor portfolios (created under (a) and (b)) correlated with each other? How strongly are the individual factor portfolios correlated with the market portfolio. What can you conclude?

```{r}
correlation_matrix <- cor(pc_scores_100, pc_scores_20)
kable(round(correlation_matrix,2))
```
The rows corresponds to the factor portfolios of the 100 size-book-to-market portfolios, and the columns to the factor portfolios of the 20 size and book-to-market portfolios. The first factors are very highly correlated with 0.99 correlation. There are also high correlations between the second factors, as well as between the third factors. PC2 and PC3 are also highly correlated between the two sets of factor portfolios.

```{r}
correlations_with_market <- sapply(colnames(pc_scores_100), function(factor) {
  cor(pc_scores_100[,factor], portfolio_m[,3])
})

kable(round(correlations_with_market,2), caption = "Correlation of each factor portfolio (of the 100 size-book-to-market) with the market portfolio:")

correlations_with_market_20 <- sapply(colnames(pc_scores_20), function(factor) {
  cor(pc_scores_20[,factor], portfolio_m[,3])
})

kable(round(correlations_with_market_20,2), caption = "Correlation of each factor portfolio (of the 20 size and book-to-market) with the market portfolio:")
```

The first factor for each of the sets are most highly correlated with the market, with the second factor having the second highest correlation with the market, and so forth. The first factors likely captures the market-wide risks and movements. Factors 3, 4, and 5 are not correlated with the market indicating that the variability these factors capture is not market wide.

## 2. Compute the excess returns for all portfolios. In the following you have to estimate a multi-factor model where you include either the principal components constructed based on the 100 size-book-to-market portfolios or the principal components constructed based on the 20 size and book-to-market portfolios as factors. Estimate the model for individual size-book-to-market portfolios in the 1st, 5th and 10th size and book-to-market decile.

```{r}
mR = as.matrix(portfolio_m[,25:124])
Rf = as.matrix(portfolio_m[,'Tbill'])
Rm = as.matrix(portfolio_m[,'Market'])
mZ = sweep(mR,1,Rf) #matrix of excess returns for the portfolios
vZm = Rm - Rf #market excess returns

pca_res_100 <- PCA(as.matrix(portfolio_m[,25:124]), 5) #5 Principal components based on the 100 sbtm portfolios
pc_scores_100 <- pca_res_100$factors
pc_scores_excess_100 <- sweep(pc_scores_100, 1, Rf) #excess returns of factor portfolios

portfolio <- c(1,5,10,41,45,50,91,95,100)
factors <- c(1,2,3,4,5)

mR2 <- matrix(NA,nrow=length(portfolio),ncol=length(factors))
mBIC <- matrix(NA,nrow=length(portfolio),ncol=length(factors))

for (i in 1:length(portfolio)){
  for (j in 1:length(factors)){
    model <- lm(mZ[, portfolio[i]] ~ pc_scores_excess_100[, 1:j])
    mR2[i, j] <- summary(model)$r.squared
    mBIC[i, j] <- BIC(model)
  }
}
```


### (a) How much of the time series variations of the portfolio returns can be explained by the individual principal components? How strong does the goodness-of-fit increase when we include more than one factor? What is the optimal number of factors according to the Bayes/Schwarz information criterion?

```{r}
R2_table <- cbind(colnames(mZ)[c(1,5,10,41,45,50,91,95,100)],round(mR2,2))
colnames(R2_table) <- c("Portfolio", "1 Factor", "2 Factors", "3 Factors", "4 Factors", "5 Factors")
kable(R2_table, caption = "R-squared values for 1-5 factor model")

BIC_table <- cbind(colnames(mZ)[c(1,5,10,41,45,50,91,95,100)],round(mBIC,0))
colnames(BIC_table) <- c("Portfolio", "1 Factor", "2 Factors", "3 Factors", "4 Factors", "5 Factors")
kable(BIC_table, caption = "BIC values for 1-5 factor models")
```

Among the first observations for the R2 we observe that the variation explained by the models increases for the portfolios in the models  1&5, 1&10, 5&1 - 5&10, 10&1, 10&5 concluding that it is harder to explain the variation in the most extreme cases 1&1 and 10&10. The other finding is that the explained variation increases as we introduce more factors, often concluding that the optimal number of factors are 4 or 5. The BIC will help us evaluate this in greater detail, especially due to the fact that BIC also penalizes a model for introducing factors that doesn't contribute to the model. 

The BIC suggests that optimal number of factors are R11: 5 , R15: 5, R110: 5, R51: 4, R55: 4, R510: 5, R101: 5, R105: 4, and R1010: 5. Please notice that the difference is very small for the portfolios where it suggests 4 instead of 5 concluding that it doesn't influence the effectiveness of the model greatly but considering that we always prefer less complex models, we should go for the 4-factor models when it is suggested. 

### (b) Do you find differences in the explanatory power of the two sets of principal components?

For simplicity lets assume 5 factor model is the best.

```{r}
pca_res_20 <- PCA(as.matrix(portfolio_m[,5:24]), 5) #20 size book, 5 factors
pc_scores_20 <- pca_res_20$factors
pc_scores_excess_20 <- sweep(pc_scores_20, 1, Rf)

pca_res_100 <- PCA(as.matrix(portfolio_m[,25:124]), 5) #100 size book, 5 factors
pc_scores_100 <- pca_res_100$factors
pc_scores_excess_100 <- sweep(pc_scores_100, 1, Rf)


portfolio <- c(1,5,10,41,45,50,91,95,100)

mR2_20_100 <- matrix(NA, nrow = length(portfolio), ncol = 2)

for (i in 1:length(portfolio)){
    model_20 <- lm(mZ[, portfolio[i]] ~ pc_scores_excess_20)
    mR2_20_100[i,1] <- summary(model_20)$r.squared
    model_100 <- lm(mZ[, portfolio[i]] ~ pc_scores_excess_100)
    mR2_20_100[i,2] <- summary(model_100)$r.squared
}


R2_comparison <- data.frame(
  Portfolio = colnames(mZ)[portfolio],
  `PCA based on 20 MEx/MEBEx portfolios` = round(mR2_20_100[, 1], 2),
  `PCA based on 100 Rx(x)y(y) portfolios` = round(mR2_20_100[, 2], 2)
)

kable(R2_comparison, caption = "R-squared values for 5 factor models")
```

Once again, we observe the for most of the cases they are quite similar. Initially and connecting to theory we would expect the $R^2$ for the 100 portfolios to be higher since this would allow for more granular analysis. For the first three portfolios, we observe that $R^2$ is considerably higher except for the R110 portfolio. In that case $R^2$is higher by 5\%. In the middlemost portfolios the difference between PCA20 and PCA100 is at most 0.01, which makes us believe there is no considerable difference in the variation explained using PCA for the 20 portfolios or 100. Lastly, when interpreting the $R^2$ for the last three rows, we observe that PCA20 would capture more variation than PCA100 except for in the last row which is R1010. This finding concludes that for the low size/book-to-market portfolios, PCA100 is preferred, in the middle it is a matter of personal preference and for the higher size/book-to-market portfolios it would be preferred using PCA20. 

We say matter of personal preference in this case since one would suggest a model with fewer factors would lesser the risk of overfitting and be more robust, however a model with a higher number of factors might capture the more granular information and estimate the broader systemic risk to a better degree. 

### (c) Evaluate the explanatory power of the individual principal components and compare it to that provided by the market risk premium solely. Does the market risk premium have any explanatory power beyond the principal components? What is about the significance of the market risk premium in such a regression? What can you conclude?

```{r}
mR = as.matrix(portfolio_m[,25:124])
Rf = as.matrix(portfolio_m[,'Tbill'])
Rm = as.matrix(portfolio_m[,'Market'])
mZ = sweep(mR,1,Rf) #matrix of excess returns for all portfolios
vZm = Rm - Rf #matrix of excess returns for the market portfolio

portfolio <- c(1,5,10,41,45,50,91,95,100)

vR2_market <- matrix(NA,nrow=length(portfolio),1)
vtstat_market <- matrix(NA,nrow=length(portfolio),1)
vpval_market <- matrix(NA,nrow=length(portfolio),1)



for (i in 1:length(portfolio)){
    model <- lm(mZ[, portfolio[i]] ~ vZm)
    vR2_market[i, 1] <- summary(model)$r.squared
    vtstat_market[i,1] <-  summary(model)$coefficients[2, "t value"]
    vpval_market[i,1] <-  summary(model)$coefficients[2, "Pr(>|t|)"]
}

R2_table <- cbind(colnames(mZ)[c(1,5,10,41,45,50,91,95,100)],round(mR2,2))
R2_table_final <- cbind(R2_table,round(vR2_market,2))
colnames(R2_table_final) <- c("Portfolio", "1 Factor", "2 Factors", "3 Factors", "4 Factors", "5 Factors", "1 Factor - Market risk premium")
kable(R2_table_final, caption = "R-squared values for 1-5 factor model and 1 Factor market model")

# Combine results into a data frame
market_stats <- data.frame(
  Portfolio = colnames(mZ)[portfolio],
  `T-statistic` = round(vtstat_market, 2),
  `P-value` = round(vpval_market, 4)
)

# Display the results as a table
kable(market_stats, caption = "Significance of the market risk premium")

```
Checking for how much of the variation is explained using the one-factor model doesn't prove any improvement compared to the 5-factor model previously used. We see that the one-factor model outperformed the other models in only one instance, for portfolio R101 and quite similar result for R105 to the result of the 5-factor model. However, it is considerably worse in all other cases, especially when we are analyzing the variation explained for low size/book-to-market portfolios. Here the model is really bad. This finding is not surprising, since empirical evidence suggests that smaller companies have a different risk-return dynamic more related to the idiosyncratic risk while the one-factor model contains the market risk premium and for this reason doesn't explain the variation that well. This might be the reasoning why Fama and French introduced additional factors size(SMB) and value(HML) to the model and developed the three-factor model. 

However, when analyzing the significance we have found that it is significant at the 1\% level for all portfolios.

### (d) For which type of portfolio do you find the best goodness-of-fit? For which type of portfolio do you find the worst goodness-of-fit? Interpret your findings.
To be able to interpret the findings we need to divide the data into three parts, small, medium and large sized companies. We observe the medium sized companies are often one of the models that exhibit the highest $R^2$ among all the models with R55 having an $R^2$ in the four and five factor models and second highest in the one-factor model. 

For small companies we observe that the ones with low and medium book-to-market ratio are experiencing consistently among the worse goodness-of-fit and this becomes especially clear in the one-factor model. However, R110 have second highest $R^2$ in the four and five factor models.

For the largest companies we observe that R101 and R105 are behaving quite similarly to the medium sized companies, experience relatively high degree of explanation and particularly R101 which is estimated with the highest $R^2$ by the one-factor model. One diverging finding here is the R1010 which is by far the absolutely worst goodness-of-fit by all the models. 

These findings can be interpreted as there are unique dynamics in the R11, R15 and R1010 which is hard to capture by the one-factor model or the PCA-models. 

Among the models we observe that the one-factor models compares generally worse than the four and five-factor models.  

## 3. Use the principal components constructed based on the 20 size and book- to-market portfolios in a multi-factor model and estimate it as a system for all 100 size-book-to- market portfolios. Test jointly for exact factor pricing in all equations. Furthermore, test for the joint significance of the factors. Use an F-test as well as a likelihood ratio test. What do you find?

We want to estimate the model

$$
Z_t=\alpha+B Z_{Kt}+\epsilon_t
$$

where $Z_{Kt}$ is the excess returns of the 5-factor portfolios with PC's based on the 20 size and book-to-market portfolios.

```{r}
mR = as.matrix(portfolio_m[,25:124])
Rf = as.matrix(portfolio_m[,'Tbill'])
mZ = sweep(mR,1,Rf) #matrix of excess returns for the portfolios

pca_res_20 <- PCA(as.matrix(portfolio_m[,5:24]), 5) #20 size book, 5 factors
pc_scores_20 <- pca_res_20$factors
mZk <- sweep(pc_scores_20, 1, Rf)
```

We can estimate the system with OLS

$$
\hat{\alpha}=\bar{Z}-\hat{\beta}\hat{\mu}_K, \\
\text{where }
\bar{Z}=\frac{1}{T} \sum_{t=1}^{T}Z_t, \text{ }\hat{\mu}_k=\frac{1}{T}\sum_{t=1}^{T}Z_{Kt},\\
\hat{\beta}= \left[\sum_{t=1}^{T}(Z_t-\bar{Z})(Z_{Kt}-\hat{\mu}_k)' \right]\left[\sum_{t=1}^{T} (Z_t-\hat{\mu}_K)(Z_{Kt}-\hat{\mu}_k)' \right]
$$
The covariance matrix of the error terms are estimated by maximum likelihood: 
$$
\hat{\Sigma}=\frac{1}{T} \sum_{t=1}^{T} (Z_t-\hat{\alpha}-\hat{\beta}Z_{Kt})(Z_t-\hat{\alpha}-\hat{\beta}Z_{Kt})'
$$

We can do the above in R:

```{r}
sur <- function(Z, Z_k){
  T <- nrow(Z)

  # Step 1: Compute means
  Z_bar <- colMeans(Z)
  mu_k <- if(!is.null(Z_k)) colMeans(Z_k) else numeric(0)  # Handle case when Z_k is NULL

  # If Z_k is NULL, estimate only intercepts
  if (is.null(Z_k)) {
    B_hat <- matrix(0, ncol(Z), 0)  
    alpha_hat <- Z_bar  
    residuals <- Z - matrix(alpha_hat, nrow = T, ncol = ncol(Z), byrow = TRUE)  
    Sigma_hat <- crossprod(residuals) / T  
  } else {
    # Step 2: Estimate B if Z_k is not NULL (i.e., perform regression)
    cov_Z_Zk <- crossprod(Z - matrix(Z_bar, nrow = T, ncol = ncol(Z), byrow = TRUE), 
                           Z_k - matrix(mu_k, nrow = T, ncol = ncol(Z_k), byrow = TRUE))
    cov_Zk <- crossprod(Z_k - matrix(mu_k, nrow = T, ncol = ncol(Z_k), byrow = TRUE))
    B_hat <- cov_Z_Zk %*% solve(cov_Zk)

    # Step 3: Estimate alpha
    alpha_hat <- Z_bar - B_hat %*% mu_k

    # Step 4: Estimate Sigma
    residuals <- Z - (matrix(alpha_hat, nrow = T, ncol = ncol(Z), byrow = TRUE) + Z_k %*% t(B_hat))
    Sigma_hat <- crossprod(residuals) / T
  }
  mE <- residuals
  ll <- LogLikelihood(mE)
  k <- dim(B_hat)[1]*dim(B_hat)[2]+length(alpha_hat)
  BIC = k*log(T)-2*ll
  
  return(list(mu_k = mu_k, B_hat = B_hat, alpha_hat = alpha_hat, Sigma_hat = Sigma_hat, residuals = residuals,BIC=BIC))
}
```

The F-test proposed by Gibbons/Ross/Shanken (1989) is:
$$
J=\frac{T-N-K}{K}[1+\hat{\mu}_K' \hat{\Omega}_K^{-1} \hat{\mu}_K]^{-1} \hat{\alpha}' \hat{\Sigma}^{-1} \hat{\alpha} \sim F_{N, T-N-K}
$$

```{r}
GRS_F <- function(Z, Z_k){
  
  sur <- sur(Z, Z_k)
  alpha_hat <- sur$alpha_hat
  mu_k <- sur$mu_k
  Sigma_hat <- sur$Sigma_hat

  N <- length(alpha_hat)
  K <- ncol(Z_k)
  T <- nrow(Z)

  Omega <- cov(Z_k)

  GRS_F <- (T - N - K) / N * 
         (t(alpha_hat) %*% solve(Sigma_hat) %*% alpha_hat) /
         (1 + t(mu_k) %*% solve(Omega) %*% (mu_k))

  GRS_p_value <- pf(GRS_F, df1 = N, df2 = T - N - K, lower.tail = FALSE)

  results <- data.frame(
    Statistic = c("GRS F-statistic:", "p-value:"),
    Value = c(GRS_F, GRS_p_value)
  )
  return(results)
}
```

And the adjusted likelihood ratio test is: 

$$
LR=(T-\frac{N}{2}-K-1) [\text{log }|\hat{\Sigma}_R|-\text{log }|\hat{\Sigma}|] \sim \chi_N^2
$$


```{r}
adjusted_LR_test <- function(sur_full, sur_res){
  T <- nrow(sur_full$residuals)
  N <- ncol(sur_full$residuals)
  K <- length(sur_full$mu_k)
  
  # Calculate LR stat, p.15 lecture 5
  adjustment_constant <- T - N / 2 - K - 1
  LR <- adjustment_constant * (logdet(sur_res$Sigma_hat) - logdet(sur_full$Sigma_hat))
  
  # p-value (chi-squared distribution)
  p_value <- pchisq(LR, df = N, lower.tail = FALSE)
  
  # Results
  results <- data.frame(
    Statistic = c("Adjusted LR statistic:", "p-value:"),
    Value = c(LR, p_value)
  )
  
  return(results)
  
}
```

Do the F-test for $\alpha=0$.
```{r, echo=F}
F_test_alpha <- GRS_F(mZ,mZk)
kable(F_test_alpha)
```

We can see that we reject the null that $\alpha=0$, so we do not have evidence of exact factor pricing. Since we reject the null that $\alpha=0$ we conclude that the factors does not achieve exact factor pricing for all pricing and there might be some systematic pricing errors in the data. 

Do the LR test for the joint significance of the factors, i.e. the do LR test with the restricted model where $B=0$

$$
Z_t=\alpha+\epsilon_t
$$

```{r, echo=F}
sur_ur_5f <- sur(mZ,mZk)
sur_r_5f <- sur(mZ,NULL)
kable(adjusted_LR_test(sur_ur_5f,sur_r_5f))
```

We can see that we reject the null that $B=0$, i.e. the factors are non-zero and jointly significant at a 1\% level. The presence of non-zero factors implies that they contribute significantly to the model of exact factor pricing. This result also validates the usage of principal components as factor in the model and is in accordance with the theory of APT which claims that systematic risks are driven by common factors. 

To summarize the results, GRS F-test rejected the null hypothesis that all intercepts are 0. In addition adj. LR test also rejected the null that the factors are zero. These two tests conclude that there are systematic pricing and that the models does not achieve exact factor pricing for all portfolios and that the non-zero factors contribute to the models significantly and are important for explaining systematic variations in portfolio returns. 


## 4. Test for joint significance of the market risk premium in all equations. Does it have explanatory power beyond the principal components? Compare the goodness-of-fit of alternative factor specifications using information criteria. Which model provides the highest goodness-of-fit for the complete system?

Now we estimate the 1 factor model

$$
Z_t=\alpha+B Z_{mt}+\epsilon_t
$$
and test if $B=0$.

```{r, echo=F}
mR = as.matrix(portfolio_m[,25:124])
Rf = as.matrix(portfolio_m[,'Tbill'])
Rm = as.matrix(portfolio_m[,'Market'])
mZ = sweep(mR,1,Rf) #matrix of excess returns for all portfolios
vZm = Rm - Rf #matrix of excess returns for the market portfolio

sur_ur_1f <- sur(mZ,vZm)
sur_r_1f <- sur(mZ,NULL)
kable(adjusted_LR_test(sur_ur_1f,sur_r_1f))

```


We reject the null that $B=0$,i .e. the market risk premium is jointly significant. Since we reject the null in both cases, the tests do not indicate that the market risk premium has explanatory power beyond the principal components.

Compare the goodness-of-fit of alternative factor specifications using information criteria. Which model provides the highest goodness-of-fit for the complete system?

```{r}
sur_ur_5f$BIC # 5 factor model with principal components
sur_ur_1f$BIC # 1 factor model with market risk premium
```
We can see that the 5 factor model with the principal components has a better goodness of fit than the 1 factor model with the market risk premium, for the complete system.


# B. Fama-French regressions
Use the data set portfolio_m and compute excess returns. In the following you have to mimic the study by Fama and French (1996) and have to investigate whether the Fama-French factors capture size and book-to-market effects.


## 1. Construct the HML and SMB factors according to Fama and French (1993) based on the return differentials of small-cap and large-cap portfolios as well as high-book-to-market and small-book-to-market portfolios.

In order to construct the HML and SMB factors, we take the mean of the three highest BtM-portfolios minus the mean of the three smallest BtM portfolios for the HML, and for SMB we take the mean of the smallest three portfolios minus the mean of the three largest portfolios.

```{r}
HML <- rowMeans(portfolio_m[,22:24])-rowMeans(portfolio_m[,15:17])
SMB <- rowMeans(portfolio_m[,5:7])-rowMeans(portfolio_m[,12:14])

portfolio_factors <- portfolio_m %>%
  mutate(HML = HML, SMB = SMB)

head(portfolio_factors[,c(1:4, 125:126)])
```

## 2. Estimate the resulting Fama-French three-factor model based on individual size-book-to-market portfolios in the 1st, 5th and 10th size and book-to-market decile.

```{r, echo=F}
cols <- c("R11", "R15", "R110", "R51", "R55", "R510", "R101", "R105", "R1010")

results <- data.frame(
  Portfolio = character(),
  Alpha = numeric(),
  Beta = numeric(),
  s = numeric(),
  h = numeric(),
  R2 = numeric(),
  R2_comp = numeric(),
  StdErrAlpha = numeric(),
  StdErrBeta = numeric(),
  StdErr_s = numeric(),
  StdErr_h = numeric(), 
  PValueAlpha = numeric(),
  PValueBeta = numeric(),
  PValue_s = numeric(),
  Pvalue_h = numeric(),
  stringsAsFactors = FALSE
)

for (i in cols){
  ER <- portfolio_factors[[i]]-portfolio_factors$Tbill
  market_excess <- portfolio_factors$Market-portfolio_factors$Tbill
  capm_model <- lm(ER ~ market_excess+portfolio_factors$SMB+portfolio_factors$HML)
  
  capm_comp <- lm(ER ~ market_excess)
  
  results <- rbind(
    results,
    data.frame(
      Portfolio = i,
      Alpha = coef(capm_model)["(Intercept)"],
      Beta = coef(capm_model)["market_excess"],
      s=coef(capm_model)["portfolio_factors$SMB"],
      h=coef(capm_model)["portfolio_factors$HML"],
      R2 = summary(capm_model)$r.squared,
      R2_comp = summary(capm_comp)$r.squared, 
      StdErrAlpha = coef(summary(capm_model))["(Intercept)", "Std. Error"],
      StdErrBeta = coef(summary(capm_model))["market_excess", "Std. Error"],
      StdErr_s = coef(summary(capm_model))["portfolio_factors$SMB", "Std. Error"],
      StdErr_h = coef(summary(capm_model))["portfolio_factors$HML", "Std. Error"], 
      PValueAlpha = coef(summary(capm_model))["(Intercept)", "Pr(>|t|)"],
      PValueBeta = coef(summary(capm_model))["market_excess", "Pr(>|t|)"],
      PValue_s = coef(summary(capm_model))["portfolio_factors$SMB", "Pr(>|t|)"],
      Pvalue_h = coef(summary(capm_model))["portfolio_factors$HML", "Pr(>|t|)"]
    )
  )
}

rownames(results) <- NULL
```

### (a) Evaluate the goodness-of-fit of the model for the individual portfolios. How much explanatory power do the HML and SMB factors have beyond the market portfolio?

```{r, echo=F}
kable(results[,c(1,6:7)], digits = 3, caption = "Comparison of R2 between the estimated three factor and CAPM model for the portfolios", label = "tab:3factor")
```
In the table above you see a comparison of the $R^2$ between the Fama and French three factor model and the regular CAPM. What we see is that the $R^2$ increases for all portfolios when including SMB and HML in the model, indicating that HML and SMB factors have explanatory power beyond the market portfolio. The increase in explanatory power is largest for the smaller portfolios (R11, R15, and R110), with the smallest increase for the largest portfolios (R101, R105, and R1010). We also see that the change is larger for the portfolios with higher book-to-market deciles (R110, R510, and R110) compared to the lower deciles. 

Based on this it seems that the explanatory power increases most for the more volatile portfolios, indicating that there is variation explained by the factors rather than the market. This reasonable as the lower book-to-market and larger size portfolios are less volatile and follow the market more closely.

### (b) Compare the estimated coefficients associated with the HML and SMB factors for portfolios in different size and book-to-market deciles. Do you find differences? Interpret your results.

```{r, echo=F}
kable(results[,c(1,4:5)], digits = 3, caption = "Comparison of estimated coefficients of the estimated three factor model between the portfolios", label = "tab:3factor_coeff")
```
Looking at the estimated coefficients above, we see that estimated $s_i$ is positive for the small and medium size portfolios (R1x and R5x), and negative for the larger size portfolios (R10x). The Estimated $s_i$ is also largest for the smallest size portfolios, and smaller for the medium and large size portfolios. The estimated $h_i$ are negative for the low book-to-market portfolios and positive for the high and medium (except R15). There is no clear correlation between the magnitude of the h coefficient and book-to-market.

Based on this we conclude that SMB more greatly explain the variation in the smaller size portfolios and that there is a negative correlation between SMB and larger size portfolios. We also conclude that HML seems to be positively correlated with high book-to-market portfolios and negatively correlated with low book-to-market portfolios.

### (c) Augment your model by the first two principal components as constructed in Problem A, Part 2. Do the principle components have explanatory power beyond the Fama- French factors? Do the Fama-French factors have explanatory power beyond the principal components? Interpret your findings.

```{r, echo=F}
colnames(pc_scores_excess_100) <- paste0("Excess_PC", 1:5)
pc_scores_excess_100 <- as.data.frame(pc_scores_excess_100)
cols <- c("R11", "R15", "R110", "R51", "R55", "R510", "R101", "R105", "R1010")

resultsb2c <- data.frame(
  Portfolio = character(),
  Alpha = numeric(),
  Beta = numeric(),
  s = numeric(),
  h = numeric(),
  PC1 = numeric(),
  PC2 = numeric(),
  R2 = numeric(),
  R2_comp = numeric(),
  StdErrAlpha = numeric(),
  StdErrBeta = numeric(),
  StdErr_s = numeric(),
  StdErr_h = numeric(), 
  StdErr_PC1 = numeric(),
  StdErr_PC2 = numeric(),
  PValueAlpha = numeric(),
  PValueBeta = numeric(),
  PValue_s = numeric(),
  Pvalue_h = numeric(),
  PValue_PC1 = numeric(),
  PValue_PC2 = numeric(),
  stringsAsFactors = FALSE
)

for (i in cols){
  ER <- portfolio_factors[[i]]-portfolio_factors$Tbill
  market_excess <- portfolio_factors$Market-portfolio_factors$Tbill
  capm_model <- lm(ER ~ market_excess + portfolio_factors$SMB + portfolio_factors$HML + pc_scores_excess_100$Excess_PC1 + pc_scores_excess_100$Excess_PC2)
  
  comp <- lm(ER ~market_excess + pc_scores_excess_100$Excess_PC1 + pc_scores_excess_100$Excess_PC2)

  resultsb2c <- rbind(
    resultsb2c,
    data.frame(
      Portfolio = i,
      Alpha = coef(capm_model)["(Intercept)"],
      Beta = coef(capm_model)["market_excess"],
      s = coef(capm_model)["portfolio_factors$SMB"],
      h= coef(capm_model)["portfolio_factors$HML"],
      PC1 = coef(capm_model)["pc_scores_excess_100$Excess_PC1"],
      PC2 = coef(capm_model)["pc_scores_excess_100$Excess_PC2"], 
      R2 = summary(capm_model)$r.squared,
      R2_comp = summary(comp)$r.squared,
      StdErrAlpha = coef(summary(capm_model))["(Intercept)", "Std. Error"],
      StdErrBeta = coef(summary(capm_model))["market_excess", "Std. Error"],
      StdErr_s = coef(summary(capm_model))["portfolio_factors$SMB", "Std. Error"],
      StdErr_h = coef(summary(capm_model))["portfolio_factors$HML", "Std. Error"], 
      StdErr_PC1 = coef(summary(capm_model))["pc_scores_excess_100$Excess_PC1", "Std. Error"],
      StdErr_PC2 = coef(summary(capm_model))["pc_scores_excess_100$Excess_PC2", "Std. Error"],
      PValueAlpha = coef(summary(capm_model))["(Intercept)", "Pr(>|t|)"],
      PValueBeta = coef(summary(capm_model))["market_excess", "Pr(>|t|)"],
      PValue_s = coef(summary(capm_model))["portfolio_factors$SMB", "Pr(>|t|)"],
      Pvalue_h = coef(summary(capm_model))["portfolio_factors$HML", "Pr(>|t|)"],
      PValue_PC1 = coef(summary(capm_model))["pc_scores_excess_100$Excess_PC1", "Pr(>|t|)"],
      PValue_PC2 = coef(summary(capm_model))["pc_scores_excess_100$Excess_PC2", "Pr(>|t|)"]
    )
  )
}

rownames(resultsb3c) <- NULL
kable(cbind(resultsb2c[,c(1,8:9)],results[,6]), digits = 3, caption = "Comparison of R2 between the estimated three factor and CAPM model for the portfolios", label = "tab:3factor_pca_r2", col.names =  c("Portfolio", "R2 (three factor + PCA)", "R2 (PCA)", "R2 (three factors)"))
```
Above you see a comparison in $R^2$ between three models: The augmented model with HML and SMB factors as well as the first two principal components, the model with only principal components, and a model with only the Fama-French factors. What we see is that including the PC's increases the goodness-of-fit compared to the three-factor model, meaning that the PC's have explanatory power beyond the Fama-French factors. However, comparing the model with only market excess and PC's we see very small differences between the $R^2$ and the augmented full model, indicating that the Fama-French factors only have small explanatory power beyond the principal components.

## 3. Estimate the Fama-French model for all 100 size-book-to-market portfolios as a system.

Based on sub-question (b), we interpret this to be the augmented model from question B.2(c). The PC's based on the 100 size-book-to-market portfolios seems to be colinear, or at least the resulting $\hat{\Omega}$ is singular such that we could not calculate the F-test. For this reason we have used the PC's computed using the 20 size and book-to-market portfolios in this question instead.

```{r, echo=F}
# Pre-process data
excess_portfolio_factors <- cbind(portfolio_factors, PC1=pc_scores_20[,1], PC2 = pc_scores_20[,2])
excess_portfolio_factors <- excess_portfolio_factors %>%
  mutate(across(c(3, 25:124,127:128), ~ . - Tbill))

# Set factors
indep <- c("Market", "SMB", "HML", "PC1", "PC2")

# As matrix
Z <- as.matrix(excess_portfolio_factors[, 25:124])
Z_k <- as.matrix(excess_portfolio_factors[, indep])

sur1 <- sur(Z, Z_k)
```

### (a) Test jointly for exact factor pricing in all equations.

We will test $H_0:\alpha=0$ using the F-test based on Gibbons/Ross/Shanken (1989):

```{r, echo=F}
results <- GRS_F(Z, Z_k)

kable(results, caption = "F-test for exact factor pricing", digits = 3)
```
Exact factor pricing implies that model explain all of the variation in excess returns, if this holds true all $\alpha$ should be zero. Based on the above F-test we reject this null hypothesis and come to the conclusion that the factor pricing model does not perfectly explain portfolio returns.

### (b) Test jointly for the significance of the HML and SMB factors as well as the principal components. What do you find? Using information criteria find the model providing the best goodness-of-fit.

For this question we perform the LR-test on slide 15 in lecture 5. We will compare the full model which includes all factors to the restricted model which only includes market excess.

```{r}
# First we estimate the restricted model which only includes "Market"
Z <- as.matrix(excess_portfolio_factors[, 25:124])
Z_k <- as.matrix(excess_portfolio_factors[, "Market"])

restricted_model <- sur(Z, Z_k)

# Compute the LR-test
adjusted_LR_results <- adjusted_LR_test(sur1, restricted_model)

kable(adjusted_LR_results, caption = "Likelihood Ratio test for significance of factors.")
```
The null hypothesis of this test is $H_0:B=0$. Based on the the low p-value of the LR test we conclude that the factors are jointly significantly different from zero.

Next, we will use BIC in order to find the model with the best goodness of fit.
```{r, echo=F}
indep1 <- c("Market")
indep2 <- c("Market", "SMB")
indep3 <- c("Market", "SMB", "HML")
indep4 <- c("PC1")
indep5 <- c("PC1", "PC2")
indep6 <- c("Market", "PC1", "PC2")
indep7 <- c("Market", "SMB", "PC1", "PC2")
indep8 <- c("Market", "SMB", "HML", "PC1", "PC2")

indep_sets <- list(indep1, indep2, indep3, indep4, indep5, indep6, indep7, indep8)

BIC_values <- sapply(indep_sets, function(indep) {
  Z_k <- as.matrix(excess_portfolio_factors[, indep])
  sur(Z, Z_k)$BIC
})

factor_names <- sapply(indep_sets, function(indep) {
  paste(indep, collapse = ", ")
})

results_df <- data.frame(
  Factors = factor_names,
  BIC = BIC_values
)

kable(results_df, caption = "BIC Values for Different Models")
```
In the above table you see the BIC for different combinations of factors which we deemed reasonable. The model with the lowest BIC is the one which includes Market, SMB, HML, PC1, and PC2, and we conclude that this model provides the best goodness of fit.


### (c) Open the data set factors_m containing the original HML and SMB factors as constructed by Fama and French (1993). Repeat the analysis above. Are your results robust regarding the exact construction of the factors?

```{r, echo=F}
SMB <- factors_m$SMB
HML <- factors_m$HML

cols <- c("R11", "R15", "R110", "R51", "R55", "R510", "R101", "R105", "R1010")

results <- data.frame(
  Portfolio = character(),
  Alpha = numeric(),
  Beta = numeric(),
  s = numeric(),
  h = numeric(),
  R2 = numeric(),
  R2_comp = numeric(),
  StdErrAlpha = numeric(),
  StdErrBeta = numeric(),
  StdErr_s = numeric(),
  StdErr_h = numeric(), 
  PValueAlpha = numeric(),
  PValueBeta = numeric(),
  PValue_s = numeric(),
  Pvalue_h = numeric(),
  stringsAsFactors = FALSE
)

for (i in cols){
  ER <- portfolio_factors[[i]]-portfolio_factors$Tbill
  market_excess <- portfolio_factors$Market-portfolio_factors$Tbill
  capm_model <- lm(ER ~ market_excess+SMB+HML)
  
  capm_comp <- lm(ER ~ market_excess+portfolio_factors$SMB+portfolio_factors$HML)
  
  results <- rbind(
    results,
    data.frame(
      Portfolio = i,
      Alpha = coef(capm_model)["(Intercept)"],
      Beta = coef(capm_model)["market_excess"],
      s=coef(capm_model)["SMB"],
      h=coef(capm_model)["HML"],
      s_comp=coef(capm_comp)["portfolio_factors$SMB"],
      h_comp=coef(capm_comp)["portfolio_factors$HML"],
      R2 = summary(capm_model)$r.squared,
      R2_comp = summary(capm_comp)$r.squared, 
      StdErrAlpha = coef(summary(capm_model))["(Intercept)", "Std. Error"],
      StdErrBeta = coef(summary(capm_model))["market_excess", "Std. Error"],
      StdErr_s = coef(summary(capm_model))["SMB", "Std. Error"],
      StdErr_h = coef(summary(capm_model))["HML", "Std. Error"], 
      PValueAlpha = coef(summary(capm_model))["(Intercept)", "Pr(>|t|)"],
      PValueBeta = coef(summary(capm_model))["market_excess", "Pr(>|t|)"],
      PValue_s = coef(summary(capm_model))["SMB", "Pr(>|t|)"],
      Pvalue_h = coef(summary(capm_model))["HML", "Pr(>|t|)"]
    )
  )
}

rownames(results) <- NULL

kable(results[,c(1,8:9)], digits = 3, caption = "Comparison of R2 between the estimated three factor models. Left: Original HML and SMB factors. Right: Our constructed factors.", label = "tab:3factor")

kable(results[,c(1,4:7)], digits = 3, caption = "Comparison of estimated coefficients of the estimated three factor models. Left: Original HML and SMB factors. Right: Our constructed factors.", label = "tab:3factor_coeff")
```
In the tables above we see that the $R^2$ are very similar between the three factor model constructed using our factors and the original Fama and French factors. There are some slight differences but based on the $R^2$, our results seems to be robust regarding the exact construction of the factors. Looking at the coefficients we see similarities between the models, there are however some differences. This is likely due to the difference in construction of the factors between ours and Fama and French's construction.

Augmenting the model in order to include the first two principal components:

```{r, echo=F}
resultsb3c <- data.frame(
  Portfolio = character(),
  Alpha = numeric(),
  Beta = numeric(),
  s = numeric(),
  h = numeric(),
  PC1 = numeric(),
  PC2 = numeric(),
  R2 = numeric(),
  R2_comp = numeric(),
  StdErrAlpha = numeric(),
  StdErrBeta = numeric(),
  StdErr_s = numeric(),
  StdErr_h = numeric(), 
  StdErr_PC1 = numeric(),
  StdErr_PC2 = numeric(),
  PValueAlpha = numeric(),
  PValueBeta = numeric(),
  PValue_s = numeric(),
  Pvalue_h = numeric(),
  PValue_PC1 = numeric(),
  PValue_PC2 = numeric(),
  stringsAsFactors = FALSE
)

for (i in cols){
  ER <- portfolio_factors[[i]]-portfolio_factors$Tbill
  market_excess <- portfolio_factors$Market-portfolio_factors$Tbill
  capm_model <- lm(ER ~ market_excess + SMB + HML + pc_scores_excess_100$Excess_PC1 + pc_scores_excess_100$Excess_PC2)
  

  resultsb3c <- rbind(
    resultsb3c,
    data.frame(
      Portfolio = i,
      Alpha = coef(capm_model)["(Intercept)"],
      Beta = coef(capm_model)["market_excess"],
      s = coef(capm_model)["SMB"],
      h= coef(capm_model)["HML"],
      PC1 = coef(capm_model)["pc_scores_excess_100$Excess_PC1"],
      PC2 = coef(capm_model)["pc_scores_excess_100$Excess_PC2"], 
      R2 = summary(capm_model)$r.squared,
      StdErrAlpha = coef(summary(capm_model))["(Intercept)", "Std. Error"],
      StdErrBeta = coef(summary(capm_model))["market_excess", "Std. Error"],
      StdErr_s = coef(summary(capm_model))["SMB", "Std. Error"],
      StdErr_h = coef(summary(capm_model))["HML", "Std. Error"], 
      StdErr_PC1 = coef(summary(capm_model))["pc_scores_excess_100$Excess_PC1", "Std. Error"],
      StdErr_PC2 = coef(summary(capm_model))["pc_scores_excess_100$Excess_PC2", "Std. Error"],
      PValueAlpha = coef(summary(capm_model))["(Intercept)", "Pr(>|t|)"],
      PValueBeta = coef(summary(capm_model))["market_excess", "Pr(>|t|)"],
      PValue_s = coef(summary(capm_model))["SMB", "Pr(>|t|)"],
      Pvalue_h = coef(summary(capm_model))["HML", "Pr(>|t|)"],
      PValue_PC1 = coef(summary(capm_model))["pc_scores_excess_100$Excess_PC1", "Pr(>|t|)"],
      PValue_PC2 = coef(summary(capm_model))["pc_scores_excess_100$Excess_PC2", "Pr(>|t|)"]
    )
  )
}

rownames(resultsb3c) <- NULL
kable(cbind(resultsb3c[,c(1,8)],resultsb2c[,8]), digits = 3, caption = "Comparison of R2 between the estimated three factor models including PC's. Left: Original HML and SMB factors. Right: Our constructed factors.", label = "tab:3factor_pca_r2", col.names =  c("Portfolio", "R2 (Original three factor + PCA)", "R2 (Our three factor + PCA)"))
```

Once again we see very similar $R^2$ between the model estimated with the original factors and our factors, indicating that our model is robust compared to the model with the original factors.

Now we will estimate the Fama-French model for all 100 size-book-to-market portfolios as a system. As in B.2(a), we will use the PC's estimated on the 20 size and book-to-market portfolios.

```{r, echo=F}
# Pre-process data
excess_portfolio_factors <- cbind(portfolio_factors, PC1=pc_scores_20[,1], PC2 = pc_scores_20[,2])
excess_portfolio_factors <- excess_portfolio_factors %>%
  mutate(across(c(3, 25:124,127:128), ~ . - Tbill))
excess_portfolio_factors$HML <- HML
excess_portfolio_factors$SMB <- SMB

# Set factors
indep <- c("Market", "SMB", "HML", "PC1", "PC2")

# As matrix
Z <- as.matrix(excess_portfolio_factors[, 25:124])
Z_k <- as.matrix(excess_portfolio_factors[, indep])

sur3c <- sur(Z, Z_k)
```

Now we test jointly for exact factor pricing in all equations. We will test $H_0:\alpha=0$ using the F-test based on Gibbons/Ross/Shanken (1989):

```{r, echo=F}
results <- GRS_F(Z, Z_k)

kable(results, caption = "F-test for exact factor pricing", digits = 3)
```
Exact factor pricing implies that model explain all of the variation in excess returns, if this holds true all $\alpha$ should be zero. Based on the above F-test we reject this null hypothesis and come to the conclusion that the factor pricing model does not perfectly explain portfolio returns.

Now w test jointly for the significance of the HML and SMB factors as well as the principal components. We use the same LR-test as in B.3(b), and we will compare the full model which includes all factors to the restricted model which only includes market excess.

```{r, echo=F}
# First we estimate the restricted model which only includes "Market"
Z <- as.matrix(excess_portfolio_factors[, 25:124])
Z_k <- as.matrix(excess_portfolio_factors[, "Market"])

restricted_model <- sur(Z, Z_k)

# Compute the LR-test
adjusted_LR_results <- adjusted_LR_test(sur3c, restricted_model)

kable(adjusted_LR_results, caption = "Likelihood Ratio test for significance of factors.")
```
The null hypothesis of this test is $H_0:B=0$. Based on the the low p-value of the LR test we conclude that the factors are jointly significantly different from zero.

Next, we will use BIC in order to find the model with the best goodness of fit.
```{r, echo=F}
indep1 <- c("Market")
indep2 <- c("Market", "SMB")
indep3 <- c("Market", "SMB", "HML")
indep4 <- c("PC1")
indep5 <- c("PC1", "PC2")
indep6 <- c("Market", "PC1", "PC2")
indep7 <- c("Market", "SMB", "PC1", "PC2")
indep8 <- c("Market", "SMB", "HML", "PC1", "PC2")

indep_sets <- list(indep1, indep2, indep3, indep4, indep5, indep6, indep7, indep8)

BIC_values <- sapply(indep_sets, function(indep) {
  Z_k <- as.matrix(excess_portfolio_factors[, indep])
  sur(Z, Z_k)$BIC
})

factor_names <- sapply(indep_sets, function(indep) {
  paste(indep, collapse = ", ")
})

results_df <- data.frame(
  Factors = factor_names,
  BIC = BIC_values
)

kable(results_df, caption = "BIC Values for Different Models")
```

The model with the lowest BIC is the one which includes Market, SMB, HML, PC1, and PC2, and we conclude that this model provides the best goodness of fit.

As the results between the models using the original Fama and French factors are almost the same as those we got using our constructed factors, we conclude that our results are robust regarding the exact construction of the factors.