---
title: "Report for PS5"
author:
- Mark Becker, Erik Lillrank & Vilim Nedic
date: ''
output:
  pdf_document:
    keep_tex: true
  df_print: kable
  html_document:
    df_print: paged
  word_document: default
header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{multirow}
  - \usepackage{caption}
  - \usepackage{subcaption}
  - \usepackage{adjustbox}
  - \usepackage{hyperref}
toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(FE)
library(forecast)
library(knitr)
library(kableExtra)
library(dplyr)
library(ggplot2)
library(psych)
library(systemfit)
library(msos)
data("portfolio_m")
```

# A. A. Estimation of GARCH processes

In this exercise, you have to model the time-varying volatility in different index series. Perform the following analysis based on daily indexes of your choice.

We use Daily S&P 500 index close price data from 1950 to 2015, obtained from the 'gt' package.

```{r}
library(gt)
close <- sp500$close #Daily S&P 500 index close from 1950 to 2015
r <- diff(log(close))
ts.plot(r)
ts.plot(r^2)
```

## 1. Analyze the dynamical properties of log returns and squared log returns. Suggest an appropriate AR-GARCH model and estimate it. Hint: you can use an R package, for example, the package ’fGarch’. Since the package ’fGarch’ allows for ARMA specification in the conditional mean, you can also fit the data by using ARMA-GARCH framework if you want.


```{r}
desc_r <- describe(r)[, c("n", "mean", "median","sd", "min", "max", "skew", "kurtosis")]
desc_r2 <- describe(r^2)[, c("n", "mean", "median","sd", "min", "max", "skew", "kurtosis")]
table <- rbind(desc_r,desc_r2)
rownames(table) <- c("r", "r^2")
kable(round(table,4))
```

```{r}
Acf(r)
Pacf(r)
Acf(r^2)
Pacf(r^2)
```
We can see that the log returns series $r$ has significant autocorrelation, so an ARMA(p,q) model for the mean seems appropriate. Also the $r^2$ seiries shows very large persistance, so a GARCH(p,q) model is appropriate in that case.

```{r}
library(fGarch)
```
